Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-05 article.Rmd
minor fixes discussioN

In Germany, institutional spending on publication fees charged by open access journals has increased over the years. These findings are consistent with the general trend of publication fees as revenue source for open access publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. They also reflect the evolving institutional support to cover these charges in Germany [@Fournier_2013]. Similar to investigations of the expenditure on publication fees at an institutional level in the UK [@Pinfield_2015], spending size varies across the German universities and research organisations included in the Open APC initiative. With a proportion of 39 % of the overall article volume, the Max Planck Society, a large non-university research organisation, supported most open access journal publications. This likely reflects the centralised library support at the Max Planck Society where the Max Planck Digital Library has managed large open access agreements with publishers over the last decade [@Schimmer_2013, @Sikora_2015]. This presumably not only resulted in a large number of supported articles, but also contributed to an advantage in monitoring and thus being able to report on this data. Many universities and research organisations, on the other hand, reported a remarkably lower number of supported articles.

To answer data curation issues while unifying UK spending data [@Neylon_2014; @woodward_2014], re-using DOIs to gather bibliographic metadata from Crossref is a promising approach. In our study, Crossref thoroughly indexed open access journal articles disclosed in the Open APC data-set, providing publisher and journal titles for 99 % of the overall article volume. Making use of metadata from Crossref, therefore, reduces the extensive validation work of bibliographic information provided that the DOI is made available in the cost data. DOI availability is a requirement needed to be addressed in other fields of the quantitative study of scholarly communication as well, such as in altmetrics to deal with heterogeneous data [@Haustein_2016]. Drawing on Crossref has the potential to increase the comparability of cost data to prepare future negotiations with publishers on open access agreements, because Crossref's metadata reflects the current landscape of academic publishing in terms of ongoing mergers of publishing houses. Future comparative studies of publication fee spending using data on the article-level also benefit from such an approach.

This study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts because the Open APC initiative uses a minimal data scheme to make self-reporting easy, and, therefore, does not track this kind of information. However, large price ranges of particular journals suggests that varying pricing schemes are in place, an observation also made earlier [@Solomon_2012, @Pinfield_2015]. Adding to this complexity, it is likely that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported to the Open APC initiative ranged between 2000 €, the DFG price cap, and 4.403 €. Although such payments made from several budgets are a proposed strategy to sustain publication funds at German universities [@Fournier_2013], these pro rata amounts cannot be identified as such within the Open APC data, leading to a possibly flawed representations of publication fee spending in Germany. In another case, one university included its contributions to the SCOAP^3 consortia and presumably divided the sum by the articles published by their authors in SCOAP^3-covered journals.^[26] This is very arbitrary, since the mean APC for an institution can only be determined after the end of a full 3-year funding cycle. Other factors affecting price variations are currency movements over the years, and different tax rates in Germany. For instance, the Max Planck Society has a limited input tax reduction. The refund of input value added tax for publication fees is 20%. Local and payment currency, as well as taxes, as suggested by @Pinfield_2015 to increase the transparency of publication fee spending, were not disclosed in the Open APC initiative's data-set at the time of our study.

It must also be noted that reporting to the Open APC initiative is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey, asking why German institutions are reluctant to share their cost data through the Open APC initiative, one institution feared that an increase in transparency would allow publishers to adjust prices in their favour. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015]. As no reliable registry of institutional open access funds or similar support structures in Germany exists, we cannot assess the number of non-participants in Germany.

Our analysis of how institutional spending per articles was distributed over publishers and journal titles indicates that open access publishing is heterogeneous and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. Ten publishers collected 92 % of open access publication fee spending, being consistent with an observed high concentration of few publishers in current academic publishing [@Larivi_re_2015]. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees as observed by @Pinfield_2015. Rather, open access publishers such as Public Library of Science (PloS), Copernicus or MPDI rank higher in our study than in the analyses of cost data in the UK.

One possible explanation why traditional publishers are less well represented in our study, is the lack of cost information about hybrid open access journals. 99 % of the overall article volume German universities and research organisations financially supported were published in fully open access journals. This presumably reflects the DFG funding policy that excludes the support of articles published in hybrid open access journals. However, while reviewing self-reported cost data from funders or countries that also support hybrid open access journals, there are smaller proportions of payment on  articles in fully open access journals. Because open access publication fee spending is fragmented and often in-transparent, it remains open to speculation whether authors affiliated with German universities and research organisations avoid opting for open access when publishing in hybrid journals, or simply use other budgets that are not covered by the Open APC initiative.

@@

Our study reveals the size and extent of spending on open access journals using publication fees in Germany. Drawing on self-reported cost data from the Open APC initiative, payments from German universities and research institutions have grown over the years. Comparing these expenditure with those from Austria and the UK, German open access funding is mostly focussed on fully open access journals, raising important questions about the role hybrid open access journals play as publication venue for researchers affiliated with the institutions included in our study. Given our findings and the general discussion on funding policies concerning hybrid open access journals, questions about if and to what extent science policy interventions, as well as the availability of institutional support, influence how researchers publish  are of particular concern.

>>>>

In Germany, institutional spending on publication fees charged by open access journals has increased over the years. These findings are consistent with the general trend of publication fees as revenue source for open access publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. They also demonstrate the growing institutional support for researchers to cover these charges in Germany [@Fournier_2013]. Similar to the expenditure on publication fees at an institutional level in the UK [@Pinfield_2015], spending size varies across the German universities and research organisations. With a proportion of 39 % of the total article volume, the Max Planck Society, a large non-university research organisation, supported most open access journal publications included in our study. A possible explanation is the centralised library support at the Max Planck Society where the Max Planck Digital Library has managed open access agreements with publishers over the last decade on behalf of most Max Planck institutes [@Schimmer_2013, @Sikora_2015]. This centralised approach presumably not only resulted in a large number of supported open access articles, but also advanced capabilities and skills needed to report on these expenditure on a regular basis. Many universities and research organisations, on the other hand, disclosed a remarkably lower number of supported articles.

To answer data curation issues while unifying UK spending data [@Neylon_2014; @woodward_2014], re-using DOIs to gather bibliographic metadata from Crossref is a promising approach. In our study, Crossref thoroughly indexed open access journal articles disclosed in the Open APC data-set, providing information about publisher and journal titles for 99 % of all articles included in the Open APC data-set. Making use of metadata from Crossref, therefore, reduces extensive validation work of bibliographic records as long as the DOIs are made available in the cost data. Beyond identifying publishers and journals, mandatory reporting of DOIs in the spending data can also increase the re-use of such data to study other aspects of open access journals using publication fees. For instance, impact analysis in the field of altmetrics are dependent on the availability of DOIs as well [@Haustein_2016]. Drawing on Crossref has the potential to increase the comparability of cost data to prepare future negotiations with publishers on open access agreements, because Crossref's metadata represents current developments in academic publishing in terms of ongoing mergers of publishing houses. In addition to these practical consequences, future comparative studies of publication fee spending using data on the article-level can also benefit from such an approach.

This study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts. The   Open APC initiative uses a minimal data scheme to incentives self-reporting, and, therefore, does not track this kind of information. However, large price ranges suggests that different pricing levels and varying pricing schemes are in place, an observation also made earlier [@Solomon_2012; @Pinfield_2015]. Adding to this complexity, it is likely that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported to the Open APC initiative ranged between 2000 €, the DFG price cap, and 4.403 €. Although such payments being made from several budgets are a proposed strategy to sustain publication funds at German universities [@Fournier_2013], these pro-rata payment were not made transparent in the Open APC data, leading to a possibly flawed representations of publication fee spending in Germany. In another case, one university included its contributions to the SCOAP^3 consortia and presumably divided the sum by the articles published by their authors in SCOAP^3-covered journals.^[26] This is very arbitrary, since averages for an institution can only be determined after the end of a full 3-year funding cycle. Other factors affecting price variations are currency movements over the years, and different tax rates in Germany. For instance, the Max Planck Society has a limited input tax reduction. The refund of input value added tax for publication fees is 20%. Local and payment currency, as well as taxes, as suggested by @Pinfield_2015 to increase the transparency of publication fee spending, were not disclosed in the Open APC initiative's data-set at the time of our study. In addition, the Open APC data-set does not track funding sources so that we cannot determine which funders co-financed publication fees.

It must also be noted that reporting to the Open APC initiative is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey, asking why German institutions are reluctant to share their cost data through the Open APC initiative, one institution feared that an increase in transparency would allow publishers to adjust prices in their favour. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015]. As no reliable registry of institutional open access funds or related support structures in Germany exists, we cannot assess the number of non-participants in Germany.

Our analysis of how institutional spending per articles was distributed over publishers and journals indicates that open access publishing is heterogeneous and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. Ten publishers collected 92 % of open access publication fee spending, being consistent with an observed high concentration of few publishers in current academic publishing [@Larivi_re_2015]. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees as observed by @Pinfield_2015. Rather, open access publishers such as Public Library of Science (PloS), Copernicus or MPDI rank higher in our study than in the analyses of cost data in the UK.

One possible explanation why traditional publishers are less well represented in our study is the lack of cost information about hybrid open access journals. 99 % of all articles German universities and research organisations financially supported were published in fully open access journals. This presumably reflects the DFG funding policy that excludes the support of articles published in hybrid open access journals. However, while reviewing self-reported cost data from Austria and the UK where hybrid open access journals are generally supported, we observe a lower proportion of payment on articles in fully open access journals. Because open access publication fee spending is fragmented and often in-transparent, it remains open to speculation whether authors affiliated with German universities and research organisations avoid opting for open access when publishing in hybrid journals, or simply use other budgets that are not covered by the Open APC initiative.

@@

Our study reveals the size and extent of spending on open access journals using publication fees in Germany. Drawing on self-reported cost data from the Open APC initiative, payments from German universities and research institutions have grown over the years. Comparing these expenditure with those from Austria and the UK, German open access funding is mostly focussed on fully open access journals, raising important questions about hybrid open access journals as publication venue. Given our findings, and in light of the general discussion about funding policies addressing hybrid open access journals, questions about whether and to what extent science policy interventions, as well as the availability of institutional support, influence how researchers publish  are of particular concern.
